Modular Model-Checking for Fault-Tolerant Distributed System Implementations

### 2016-12-04 doc/om-paper/paper.tex
started paper edits...

\lee{grep to get consistent manifest/benign, also lieutenant vs relay/receiver.}

@@

The verification of this class of systems has usually been at the algorithmic level, eliding details about a concrete implementation, and historically, has relied on formal models verified by interactive theorem-proving~\cite{om-acl2-impl,Young97:IC,csl-93-2,pvs}. If formal verification is to be introduced into the workflow of system designers, though, we need more automated methods. (Mostly) automated proof techniques are required to reduce the need for specialized verification expertise. We also need programmatic verification of \emph{implementations}. System designers create software and hardware implementations to test, simulate, and deploy. Discrepancies between implementations and algorithmic models can arise if the latter is abstracted too much from the former, particularly if those abstractions are ad-hoc and system specific \ben{for example?}. Furthermore, as implementations are modified to explore the design space, it is easy for the formal model and the implementation to become inconsistent.

@@

In this paper, we develop a formal model to address these concerns. We combine ideas from the literature and folklore to build scalable and modular formal models of fault-tolerant distributed systems. We describe a generic formal model for fault-tolerant distributed systems that aims to reduce the need for ad-hoc verification abstractions and optimizations while maintaining scalability in Section~\ref{sec:model}. There are three aspects of the model: (1) \emph{calendar automata}, originally developed by Dutertre and Sorea~\cite{cal} for specifying real-time systems; (2) \emph{synchronous observers}, originally from the synchronous languages community~\cite{Halbwachs-Lagnier:92,Halbwachs-Lagnier:94} and popularized by Rushby for model-checking~\cite{Rushby:SAS14}; and (3) \emph{abstract state machines}, useful for abstracting counterexamples. The model is suitable for infinite-state model-checking.

To demonstrate the modeling technique, in Section~\ref{sec:byz} we present a formally-verified implementation of the Oral Messages Hybrid algorithm (\OMH)~\cite{Lincoln-Rushby}, implementing the Boyer-Moore Fast Majority Vote algorithm for algorithm~\cite{Boyer-Moore:91}. The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm. In Section~\ref{sec:experimental}, we give performance metrics for the model. First, we benchmark the model, showing that despite the additional complexity, it scales farther than other model-checking verifications at the algorithmic level. The scalability is due principally to specifying invariants and using $k$-induction in SRI's SAL model-checker~\cite{sal}. Developing invariants requires some user guidance. A user should not have to discover all new invariants for new protocols or implementations. We show that most of the lemmas required are not protocol specific and are modular, concerning only specific aspects of the system. To demonstrate this, we modify the OHM implementation in various ways to show that only local modifications are required to the model and proofs.

@@

% ------------------------------------------------------------
\section{Formal Model}\label{sec:model}
Here we describe our formal model specialized for fault-tolerant distributed systems. The model draws on three principal abstractions: calendar automata, synchronous observers, and abstract state machines; we describe each below.

@@

However, real-time system verification in general-purpose model-checkers requires an explicit formalism of real-time progression. Trying to encode real-time clocks directly is difficult; in particular, one must avoid Zeno's paradox in which no progress is made because state transitions simply update real-valued variables by an infinitesimally small amount~\cite{bruno,lamport}. To avoid this problem, Dutetre and Sorea developed \emph{calendar automata}~\cite{bruno}, which is itself inspired by event calendars used in discrete-event simulation.

@@

\item \label{cal:d} In all states, if the current time equals a timeout, then the only transitions enabled are calendar event updates associated with the timeout: $\forall \sigma \in S$, $\exists (e_i, t_i) \in \sigma(cal)$ such that $\sigma(now) = t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma'(now) = \sigma(now)$, $\sigma'(c_j) = \sigma(c_j)$ for all $c_j \in \sigma(c)$ such that $c_j \neq c_i$, and $\sigma'(t_i) > \sigma(t_i)$.
\end{enumerate}
\ben{Do we also require that $c_i \notin \sigma'(c)$ ?}

@@

\lee{technically, atomic time is just a convenience.}
The event of receiving a message initiates a process to update its local state machine and generate additional messages to send. When the process is updating its local state machine, the event calendar is paused. That is, updating an event $recv(i, j, t, m)$ also includes updating $j$'s state machine.

\lee{note that we don't need full generality of send and receive events in OM(1)}

\lee{note that we use calendar automata for non real-time system}

\subsection{Synchronous Observers}\label{sec:sync}

\lee{Lee will rework this section; update references to it.}

Synchronous observers are state machines that are synchronously composed with a system specification to check properties over the system's state variables, i.e. they are a method for composing specifications with a program. Synchronous observers were first introduced in the context of the synchronous programming languages, such as Lustre~\cite{lustre:ieee}. While synchronous observers have been part of model-checking folklore for some time, Rushby systematically describes their application to the present domain in~\cite{Rushby:SAS14}. In Section~\ref{sec:sync-model} we define synchronous observers and then describe their application to modeling faults in Section~\ref{sec:kibitzer}.

\lee{what we have defined are actually strictly stronger than synchronous observers since they can modify state.}

\subsubsection{Model}
A synchronous observer is of the form
$$\mathcal{S} | \mathcal{O}$$
\noindent
where `$|$' denotes synchronous composition of transition systems (defined formally below). $\mathcal{S}$ is the \emph{system}, and $\mathcal{O}$ is the \emph{observer}. We define synchronous composition more precisely below, beginning with some preliminary definitions:

\begin{definition}[Consistent states]
  States $s, s'$ are consistent iff for every assignment $(x = v) \in s$ and $(x' =
  v') \in s'$, $x \neq x'$ or $v = v'$.
\end{definition}

\begin{definition}[Reachable states]
For a transition relation $\rightarrow$ over the states of $S$, for states $s_0,  s \in S$, we say $s$ is reachable from $s_0$ iff there exist states $s_1, \ldots, s_k \in S$ such that
$$s_0 \rightarrow s_1 \rightarrow \ldots \rightarrow s_k \rightarrow s$$
\end{definition}
\noindent
By convention, every state $s$ is reachable from itself.

In the following definitions, let
$M = (S, I, \rightarrow)$
and
$M' = (S', I', \rightarrow')$
be transitions systems.

%% \begin{definition}[Transition system union]
%% %% $M \cup M' = (S \cup S', I \cup I', \rightarrow_{M|M'})$ where $\rightarrow_{M|M'}$ is the smallest relation such that $(s,s') \in \rightarrow_{M|M'}$ iff 
%% \end{definition}

In defining the synchronous composition of two transition systems, we allow the possibility that composing two states can lead to inconsistent states (as opposed to the semantics in which composition of states $s_0$ and $s_1$ is the ordered pair $(s_0, s_1)$).

\begin{definition}[Synchronous composition]
The synchronous composition of $M$ and $M'$ is a transition system
$$M | M' = (S_{M | M'}, I_{M | M'}, \rightarrow_{M | M'})$$
\noindent
such that
\begin{itemize}
\item $S_{M | M'} = S \cup S'$,
\item $I_{M | M'} = \{ s_I \cup s_{I'} | s_I \in I\textnormal{, }s_{I'} \in I'\textnormal{, and } s_I \textnormal{ and } s_{I'}\textnormal{ are consistent}\}$
\item and for $s, s' \in S_{M | M'}$, $s \rightarrow_{M | M'} s'$ iff
  \begin{itemize}
  \item there exist states $s_M, s'_M \in S$ such that $s_M \rightarrow s'_M$, $s_M \subseteq s$, and $s'_M \subseteq s'$,
  \item there exist states $s_{M'}, s'_{M'} \in S'$ such that $s_{M'} \rightarrow s'_{M'}$, $s_{M'} \subseteq s$, and $s'_{M'} \subseteq s'$,
  \item $s_M$ and $s_{M'}$ are consistent,
  \item and $s'_M$ and $s'_{M'}$ are consistent.
  \end{itemize}
\end{itemize}
\end{definition}

@@

There are two practical benefits of the synchronous observer approach. First, the system $\mathcal{S}$ contains no extraneous modeling context about the environment, assumptions, or requirements, making it easier to decompose $\mathcal{S}$ and to synthesize an implementation directly from it. Second, as described by Rushby~\cite{}, requirements can be specified using transition systems rather than temporal logic formula, which is often more natural. We use synchronous observers for both purposes.

\subsubsection{The Synchronous Kibitzer}\label{sec:kibitzer}
Synchronous observers can be used to model the environment, and in particular, inject faults in a compositional way. The typical approach to injecting faults into a formal model of a distributed system is to add new state variables to each process representing its fault state. Then additional state variables are added to represent the non-determinism introduced by faults into the state or messages by a process.

\lee{cite abstractions paper and point out that we can model faults on transmission rather than node}

@@

Not only is modeling faults generally expensive, but in typical practice it is generally non-compositional insofar as the fault model pervades the entire model.

Another difficulty with adding faults to a model is that the number of states that must be introduced may depend non-obviously on other aspects of the fault model, specific protocol, and system size. Furthermore, since SAT-based model-checking is NP-complete, it is important not to introduce any state beyond the minimal required. Such constraints lead to ``meta-model'' reasoning, such as the following, as described by Rushby:

\begin{quote}
To achieve the full range of faulty behaviors, it seems that a faulty source should be able to send a \emph{different} incorrect value to each relay, and this requires $n$ different values. It might seem that we need some additional incorrect values so that faulty relays can exhibit their full range of behaviors. It would certainly be safe to introduce additional values for this purpose, but the performance of model checking is very sensitive to the size of the state space, so there is a countervailing argument against introducing additional values. A little thought will show that the way faulty relays have their most significant impact is by tipping the majority vote in a receiver one way or the other, and this can only be achieved if they use the same values as nonfaulty relays. Hence, we decide against further extension to the range of values~\cite{Rushby:OM1}.
\end{quote}

@@

\subsection{Model Sketch}
The verification is carried out in the Symbolic Analysis Laboratory~\cite{}, which contains a suite of model-checkers and a high-level specification language. In our work, we use infinite-state (SMT-based) $k$-induction~\cite{}.

>>>>

\lee{grep to get consistent manifest/benign, also lieutenant vs relay/receiver.}
\lee{grep for synchronous observers section references}

@@

The verification of this class of systems has usually been at the algorithmic level, eliding details about a concrete implementation, and historically, has relied on formal models verified by interactive theorem-proving~\cite{om-acl2-impl,Young97:IC,csl-93-2,pvs}. If formal verification is to be introduced into the workflow of system designers, though, we need more automated methods. (Mostly) automated proof techniques are required to reduce the need for specialized verification expertise. We also need programmatic verification of \emph{implementations}. System designers create software and hardware implementations to test, simulate, and deploy. Discrepancies between implementations and algorithmic models can arise if the latter is abstracted too much from the former~\cite{paxos}, particularly if those abstractions are ad-hoc and system specific. Furthermore, as implementations are modified to explore the design space, it is easy for the formal model and the implementation to become inconsistent.

@@

In this paper, we develop a formal framework to address these concerns. We combine ideas from the literature and folklore to build scalable and modular formal models of fault-tolerant distributed systems. We describe a generic formal framework for fault-tolerant distributed systems that aims to reduce the need for ad-hoc verification abstractions and optimizations while maintaining scalability in Section~\ref{sec:model}. There are three aspects of the framework: (1) \emph{calendar automata}, originally developed by Dutertre and Sorea~\cite{cal} for specifying real-time systems; (2) \emph{synchronous observers}, originally from the synchronous languages community~\cite{Halbwachs-Lagnier:92,Halbwachs-Lagnier:94} and popularized by Rushby for model-checking~\cite{Rushby:SAS14}; and (3) \emph{abstract state machines}, useful for abstracting counterexamples. The framework is suitable for infinite-state model-checking.

To demonstrate the framework, in Section~\ref{sec:byz} we present a formally-verified implementation of the Oral Messages Hybrid algorithm (\OMH)~\cite{Lincoln-Rushby}, implementing the Boyer-Moore Fast Majority Vote algorithm for algorithm~\cite{Boyer-Moore:91}. The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm. In Section~\ref{sec:experimental}, we give performance metrics for the model. First, we benchmark the model, showing that despite the additional complexity, it scales farther than other model-checking verifications at the algorithmic level. The scalability is due principally to specifying invariants and using $k$-induction in SRI's SAL model-checker~\cite{sal}. Developing invariants requires some user guidance. A user should not have to discover all new invariants for new protocols or implementations. We show that most of the lemmas required are not protocol specific and are modular, concerning only specific aspects of the system. To demonstrate this, we modify the OHM implementation in various ways to show that only local modifications are required to the model and proofs.

@@

% ------------------------------------------------------------
\section{Formal Framework}\label{sec:model}
Here we describe our formal framework specialized for fault-tolerant distributed systems. The framework draws on three principal abstractions: calendar automata, synchronous observers, and abstract state machines; we describe each below.

@@

However, real-time system verification in general-purpose model-checkers requires an explicit formalism of real-time progression. Trying to encode real-time clocks directly is difficult; in particular, one must avoid Zeno's paradox in which no progress is made because state transitions simply update real-valued variables by an infinitesimally small amount~\cite{bruno,lamport}. To avoid this problem, Dutetre and Sorea developed \emph{calendar automata}~\cite{bruno}, which is itself inspired by event calendars used in discrete-event simulation. Indeed, while calendar automata have historically been used for verifying real-time systems, they can be used to verify non real-time systems, too, specified in a discrete-event formulation. We discuss taking a non real-time model and easily transforming it to a real-time time-triggered model in Section~\ref{sec:modular}.

@@

\item \label{cal:d} In all states, if the current time equals a timeout, then the only transitions enabled are calendar event updates associated with the timeout: $\forall \sigma \in S$, $\exists (e_i, t_i) \in \sigma(cal)$ such that $\sigma(now) = t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma'(now) = \sigma(now)$, $\sigma'(c_j) = \sigma(c_j)$ for all $c_j \in \sigma(c)$ such that $c_j \neq c_i$, and $c_i \notin \sigma'(c)$.

\end{enumerate}

@@

The event of receiving a message initiates a process to update its local state machine and generate additional messages to send. When the process is updating its local state machine, the event calendar is paused. That is, updating an event $recv(i, j, t, m)$ also includes updating $j$'s state machine.

%% \subsection{Synchronous Observers and }\label{sec:sync}

\lee{Cutting synchronous observer section; need to talk about use for lemmas elsewhere.}

%% Synchronous observers are state machines that are synchronously composed with a system specification to check properties over the system's state variables, i.e. they are a method for composing specifications with a program. Synchronous observers were first introduced in the context of the synchronous programming languages, such as Lustre~\cite{lustre:ieee}. While synchronous observers have been part of model-checking folklore for some time, Rushby systematically describes their application to the present domain in~\cite{Rushby:SAS14}. In Section~\ref{sec:sync-model} we define synchronous observers and then describe their application to modeling faults in Section~\ref{sec:kibitzer}.

%% \lee{what we have defined are actually strictly stronger than synchronous observers since they can modify state.}

%% \subsubsection{Model}
%% A synchronous observer is of the form
%% $$\mathcal{S} | \mathcal{O}$$
%% \noindent
%% where `$|$' denotes synchronous composition of transition systems (defined formally below). $\mathcal{S}$ is the \emph{system}, and $\mathcal{O}$ is the \emph{observer}. We define synchronous composition more precisely below, beginning with some preliminary definitions:

%% \begin{definition}[Consistent states]
%%   States $s, s'$ are consistent iff for every assignment $(x = v) \in s$ and $(x' =
%%   v') \in s'$, $x \neq x'$ or $v = v'$.
%% \end{definition}

%% \begin{definition}[Reachable states]
%% For a transition relation $\rightarrow$ over the states of $S$, for states $s_0,  s \in S$, we say $s$ is reachable from $s_0$ iff there exist states $s_1, \ldots, s_k \in S$ such that
%% $$s_0 \rightarrow s_1 \rightarrow \ldots \rightarrow s_k \rightarrow s$$
%% \end{definition}
%% \noindent
%% By convention, every state $s$ is reachable from itself.

%% In the following definitions, let
%% $M = (S, I, \rightarrow)$
%% and
%% $M' = (S', I', \rightarrow')$
%% be transitions systems.

%% %% \begin{definition}[Transition system union]
%% %% %% $M \cup M' = (S \cup S', I \cup I', \rightarrow_{M|M'})$ where $\rightarrow_{M|M'}$ is the smallest relation such that $(s,s') \in \rightarrow_{M|M'}$ iff 
%% %% \end{definition}

%% In defining the synchronous composition of two transition systems, we allow the possibility that composing two states can lead to inconsistent states (as opposed to the semantics in which composition of states $s_0$ and $s_1$ is the ordered pair $(s_0, s_1)$).

%% \begin{definition}[Synchronous composition]
%% The synchronous composition of $M$ and $M'$ is a transition system
%% $$M | M' = (S_{M | M'}, I_{M | M'}, \rightarrow_{M | M'})$$
%% \noindent
%% such that
%% \begin{itemize}
%% \item $S_{M | M'} = S \cup S'$,
%% \item $I_{M | M'} = \{ s_I \cup s_{I'} | s_I \in I\textnormal{, }s_{I'} \in I'\textnormal{, and } s_I \textnormal{ and } s_{I'}\textnormal{ are consistent}\}$
%% \item and for $s, s' \in S_{M | M'}$, $s \rightarrow_{M | M'} s'$ iff
%%   \begin{itemize}
%%   \item there exist states $s_M, s'_M \in S$ such that $s_M \rightarrow s'_M$, $s_M \subseteq s$, and $s'_M \subseteq s'$,
%%   \item there exist states $s_{M'}, s'_{M'} \in S'$ such that $s_{M'} \rightarrow s'_{M'}$, $s_{M'} \subseteq s$, and $s'_{M'} \subseteq s'$,
%%   \item $s_M$ and $s_{M'}$ are consistent,
%%   \item and $s'_M$ and $s'_{M'}$ are consistent.
%%   \end{itemize}
%% \end{itemize}
%% \end{definition}

@@

%% There are two practical benefits of the synchronous observer approach. First, the system $\mathcal{S}$ contains no extraneous modeling context about the environment, assumptions, or requirements, making it easier to decompose $\mathcal{S}$ and to synthesize an implementation directly from it. Second, as described by Rushby~\cite{}, requirements can be specified using transition systems rather than temporal logic formula, which is often more natural. We use synchronous observers for both purposes.

\subsubsection{A Synchronous Kibitzer}\label{sec:kibitzer}
The typical approach to modeling faults is to add new state variables to each process representing its fault state. Then additional state variables are added to represent the non-determinism introduced by faults into the state or messages by a process. As a simple example, we might define a node that sends a good message if it is non-faulty and a bad message otherwise. In pseudo-code using guarded commands, its definition might look like the following:

\begin{verbatim}
node:
  health: Fault\_Type;

  faulty(health)      --> send(bad\_msg);
  non\_faulty(health) --> send(good\_msg);
\end{verbatim}

@@

\noindent
But this typical approach mixes the specification of a node's behavior with the fault model, an aspect of the environment. Doing so divorces a node specification from its implementation.

A second difficulty with model-checking fault-tolerant systems in general is that modeling faults requires adding state and nondeterminism. The minimum number of additional states that must be introduced may depend non-obviously on other aspects of the fault model, specific protocol, and system size. Such constraints lead to ``meta-model'' reasoning, such as the following, as described by Rushby:

\begin{quote}
To achieve the full range of faulty behaviors, it seems that a faulty source should be able to send a \emph{different} incorrect value to each relay, and this requires $n$ different values. It might seem that we need some additional incorrect values so that faulty relays can exhibit their full range of behaviors. It would certainly be safe to introduce additional values for this purpose, but the performance of model checking is very sensitive to the size of the state space, so there is a countervailing argument against introducing additional values. A little thought will show that $\ldots$. Hence, we decide against further extension to the range of values~\cite{Rushby:OM1}.
\end{quote}

@@

\subsection{Model Sketch}\label{sec:sketch}
\lee{Ben: can you introduce the relay/receiver unrolling of lieutenants here, since I refer to them later?}

The verification is carried out in the Symbolic Analysis Laboratory~\cite{}, which contains a suite of model-checkers and a high-level specification language. In our work, we use infinite-state (SMT-based) $k$-induction~\cite{}.
